Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2023-07-20 ba_resubmit/kundinger_final.tex
done with comments in 4.2

After computing all comparison vectors within a block, the modeler can further reduce the number of comparison vectors used in the linkage algorithm through indexing. For example, one might only consider pairs with a certain similarity score on a field deemed to be important, like last name, or only pairs that exactly match on a specified number of fields. However, the impact of indexing on model parameters is not well understood; \citep{murray2016probabilistic} reviewed this issue in the context of the frequentist Fellegi-Sunter model, leaving the effect of indexing on Bayesian record linkage models to future work.

@@

Since each component $\gamma_{ij}^f$ is discrete, there are only finitely many possible realizations of the comparison vector $\gamma_{ij}$. Let $P$ be the number of patterns realized in $\gamma$. 
%It is
\textcolor{teal}{The comparison vector} is
 bounded above by $P^{*} =  \prod_{f=1}^F (L_f + 1).$
% This comment below seems obvious and not needed.  
% where the addition of 1 to $L_f$ for each field accounts for the possibility of missing values. 
\textcolor{teal}{The quantity $P^{*}$ clearly depends only on $F$ and $L_f.$ Observe it does not depend (or scale) with $n_A$ or $n_B.$}
%is determined by $F$ and $L_f$, and does not scale with $n_A$ or $n_B$.

@@

%\textcolor{red}{Do you mean what I have below or something else? This still is a bit hard to parse.}
We then map \textcolor{teal}{these integers in (\ref{eqn:hashing})} to sequential integers $\{1, \ldots, P\}$. \textcolor{teal}{Denote} each unique agreement pattern as $h_p$, and the set of unique agreement patterns as $\mathcal{P} = \{h_1, \ldots, h_P\}$. When record pair, \textcolor{teal}{$(i,j)$,} exhibits agreement pattern $p$, \textcolor{teal}{let} $\gamma_{ij} = h_p$.

@@

\begin{example}
\label{ex:one-hot}
For example, with five fields with binary agreements and possible missingness (NA), the number of possible patterns is bounded above by $P^{*} = 3^5 = 243$. Records $A_5$ and $B_7$ may exhibit agreement pattern $\gamma_{5,7} = (1, 1, 1, \text{NA}, 2)$, indicating exact agreement on the first three fields, missing information in the fourth field, and disagreement in the fifth field. Then (\ref{eqn:hashing}) gives $h^{*}(\gamma_{5,7}) = 2^1 + 2^3 + 2^5 + 0 + 2^{10} = 1066$. We map the unique values of $h^{*}(\gamma)$ to sequential integers. \textcolor{teal}{Suppose that 1066 maps to 42. This implies $\gamma_{5,7} = h_{42}.$}
%If 1066 is mapped to, for example, 42, then we would have $\gamma_{5,7} = h_{42}.$ 
This agreement pattern has the one hot encoding $e(h_{42}) = (1, 0, 1, 0, 1, 0, 0, 0, 0, 1)$.
\end{example}

We then identify the records in $A$ with comparison vectors corresponding to each pattern $p$ for each record $B_j$. We denote this set $r_{p_j} = \{i \in 1, \dots, n_A | \gamma_{ij} = h_p\}$, and collect all such sets in the nested list $\mathcal{R} = \{r_{p_j} | p \in \{1, \ldots, P\}, j \in \{1, \ldots, n_B\} \}$. We compute the number of records in $A$ that share agreement pattern $p$ with record $B_j$, given by
\begin{align}\label{eqn:N}
N_{p_j} = |r_{p_j}| = \sum_{i=1}^{n_A} I(\gamma_{ij} = h_p).
\end{align}
We collect these counts in $\mathcal{N} = \{N_{p_j} |p \in 1, \ldots, P, j \in 1, \ldots, n_B \}$.

Together, the set $\tilde{\gamma} = \{\mathcal{P}, \mathcal{R}, \mathcal{N}\}$ fully characterizes the comparison matrix $\gamma$ for purposes of writing the likelihood function for \texttt{fabl}. To see this, we employ the condensed notation
\begin{subequations}
	\begin{align}
		m_p =  p(\gamma_{ij} = h_p|Z_j = i) = \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}\label{eqn:m_p} \\
		u_p =  p(\gamma_{ij} = h_p|Z_j \neq i) = \prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}\label{eqn:u_p}
	\end{align}
\end{subequations}
to express the probability that records $A_i$ and $B_j$ form agreement pattern $p$ given that they are a match, in (\ref{eqn:m_p}), and not a match, in (\ref{eqn:u_p}). Viewed through the perspective of agreement patterns, the likelihood in (\ref{eqn:likelihood}) \textcolor{teal}{can be expressed as}
%is equivalent to

@@

Turning first to the \textcolor{teal}{full conditional updates}
%posterior updates 
for $\bm{m}$ and $\bm{u}$, we observe that these parameters only depend on the \textcolor{teal}{observed} data  through quantities that are calculated through $\mathcal{N}$ and $\mathcal{P}$. Let $n_p(\bm{Z}) = \sum_{j=1}^{n_B} I\left(\gamma_{Z_j, j} = h_p \right)$ be the number of matching record pairs with agreement pattern $p$, and $N_p = \sum_{j=1}^{n_B} N_{p_j}$ be the total occurrence of pattern $p$ in the data across all record pairs. Then, conditional on $Z$, we can express the contribution to the likelihood in the full conditional distribution for $\bm{m}$ and $\bm{u}$ as
\begin{align}
	\mathcal{L}(\bm{m}, \bm{u} \mid \tilde{\gamma}, \bm{Z}, \pi)  &=  \prod_{p=1}^P m_p^{n_p(\bm{Z})}u_p^{N_p - n_p(\bm{Z})}  \label{likelihood_efficient_m_u}.
\end{align}
Additionally, let $\bm{\alpha_0} = (\alpha_{11}, \ldots, \alpha_{F L_F})$ and $\bm{\beta_0}= (\beta_{11}, \ldots, \beta_{F L_F})$ be a concatenated vectors of prior parameters for the $\bm{m}$ and $\bm{u}$ distributions respectively. The terms needed for the \textcolor{teal}{full conditional} updates for the $\bm{m}$ and $\bm{u}$ parameters are given by the appropriate components of the vectors
\begin{subequations}
	\begin{align}
		\bm{\alpha(Z}^{(s)}\bm{)} &= \bm{\alpha_0} + \sum_{p=1}^P n_p\left(\bm{Z}^{(s)}\right) \times e(h_p), \label{efficient_alpha} \\
		\bm{\beta(Z}^{(s)}\bm{)} &= \bm{\beta_0} + \sum_{p=1}^P \left(N_p - n_p\left(\bm{Z}^{(s)}\right)\right) \times e(h_p). \label{efficient_beta}
	\end{align}
\end{subequations}
%Specifically, the $\ell + \sum_{k=1}^{f-1} L_k$ component of (\ref{efficient_alpha}) and (\ref{efficient_beta}) provides the the posterior updates for level $\ell$ and field $f$ in (\ref{eqn:alpha_update}) and (\ref{eqn:beta_update}). However, the vectorized summations in (\ref{efficient_alpha}) and (\ref{efficient_beta}) can be computed more efficiently than summing over $n_A n_B$ record pairs for each field and agreement level.

@@

\textcolor{teal}{Observe that we compute the vectorized summations in (\ref{efficient_alpha}) and (\ref{efficient_beta}) more efficiently than the previous updates in (\ref{eqn:alpha_update}) and (\ref{eqn:beta_update}), which require summing over $n_A n_B$ record pairs for each field and agreement level.}

\textcolor{teal}{Turning to sampling for $\pi,$ it remains the same as in (\ref{eqn:pi}). }

We also use (\ref{likelihood_efficient}) to sample $\bm{Z}$ efficiently. For each pattern $p$, let $w_p =\frac{m_p}{u_p}$.  The contribution to the likelihood in the full conditional for $Z_j$ can be expressed as
\begin{align}
	\mathcal{L}(Z_j\mid  \tilde{\gamma}, \bm{m}, \bm{u}, \pi) = \prod_{p=1}^P u_p^{N_{p_j}} \prod_{i \in r_{p_j}} w_p^{I(Z_j = i)}. \label{likelihood_efficient_z}
\end{align}
This likelihood lends itself to markedly faster posterior computations. Sampling $Z_j$ from the full conditional provided in (\ref{eqn:z_full_conditional}) can become computationally expensive when $n_A$ is large. This is because sampling a value from $n_A$ options with unequal weights requires normalizing the weights to probabilities, which has a computational cost that scales with $n_A$. \textcolor{teal}{We consider two sampling steps.}
%To speed up computation, we break this sampling step into two.

We first sample among $P + 1$ options for the agreement pattern between $B_j$ and its potential link, according to
\begin{align}
	\label{eqn:gibbs1}
	P\left( \gamma_{Z_j^{(s+1)}, j} = h_p \mid \tilde{\gamma}, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)}\right) \propto
	\begin{cases} 
		\frac{\pi^{(s+1)}N_{p_j}}{n_A}  w_{p}^{(s+1)},  & \gamma_{Z_j^{(s+1)}, j} = h_p; \\
		1- \pi^{(s+1)} , &   \text{otherwise}. \\
	\end{cases}
\end{align}
Since all records in $A$ sharing the same agreement pattern with $B_j$ are equally likely, we then sample among candidate records uniformly using
\begin{align}
	\label{eqn:gibbs2}
	P\left(Z_j^{(s+1)} = q \mid \gamma_{Z_j^{(s+1)}, j} = h_p, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)} \right) = \begin{cases} 
		\frac{1}{N_{p_j}}, & q \in r_{p_j}; \\
		0, & \text{otherwise.} \\
	\end{cases}
\end{align}

>>>>

After computing all comparison vectors within a block, the modeler can further reduce the number of comparison vectors used in the linkage algorithm through indexing. For example, one might only consider pairs with a certain similarity score on a field deemed to be important, like last name, or only pairs that exactly match on a specified number of fields. However, the impact of indexing on model parameters is not well understood; \cite{murray2016probabilistic} reviewed this issue in the context of the \textcolor{red}{classical} Fellegi-Sunter model, leaving the effect of indexing on Bayesian record linkage models to future work.

@@

Since each component $\gamma_{ij}^f$ is discrete, there are only finitely many possible realizations of the comparison vector $\gamma_{ij}$. Let $P$ be the number of patterns realized in $\gamma$. 
\textcolor{red}{The quantity $P$ is
 bounded above by $P^{*} =  \prod_{f=1}^F (L_f + 1),$ which clearly depends only on $F$ and $L_f.$ It does not depend (or scale with) with $n_A$ or $n_B.$}
% This comment below seems obvious and not needed.  
% where the addition of 1 to $L_f$ for each field accounts for the possibility of missing values.

%is determined by $F$ and $L_f$, and does not scale with $n_A$ or $n_B$.

@@

%\textcolor{red}{Do you mean what I have below or something else? This still is a bit hard to parse.}
We then map \textcolor{teal}{the integers in (\ref{eqn:hashing})} to sequential integers $\{1, \ldots, P\}$. \textcolor{teal}{Denote} each unique agreement pattern as $h_p$, and the set of unique agreement patterns as $\mathcal{P} = \{h_1, \ldots, h_P\}$. When record pair, \textcolor{teal}{$(i,j)$,} exhibits agreement pattern $p$, \textcolor{teal}{let} $\gamma_{ij} = h_p$.

@@

\begin{example}
\label{ex:one-hot}
%% Jerry propose with (want to avoid using with twice in the same sentence :)
\textcolor{teal}{Consider} five fields with binary agreements and possible missingness (NA). The number of possible patterns is bounded above by $P^{*} = 3^5 = 243$. \textcolor{red}{Suppose that} records $A_5$ and $B_7$ exhibit agreement pattern $\gamma_{5,7} = (1, 1, 1, \text{NA}, 2)$, indicating exact agreement on the first three fields, missing information in the fourth field, and disagreement in the fifth field. Then (\ref{eqn:hashing}) gives $h^{*}(\gamma_{5,7}) = 2^1 + 2^3 + 2^5 + 0 + 2^{10} = 1066$. 
\textcolor{red}{All records with a hashed value of 1066 map to the integer 42. Then, $\gamma_{5,7} = h_{42}.$}
%We map the unique values of $h^{*}(\gamma)$ to sequential integers. \textcolor{teal}{Suppose that 1066 maps to 42. This implies $\gamma_{5,7} = h_{42}.$}
%If 1066 is mapped to, for example, 42, then we would have $\gamma_{5,7} = h_{42}.$ 
This agreement pattern has the one hot encoding $e(h_{42}) = (1, 0, 1, 0, 1, 0, 0, 0, 0, 1)$.
\end{example}

We then identify the records in $A$ with comparison vectors corresponding to each pattern $p$ for each record $B_j$. We denote this set $r_{p_j} = \{i \in 1, \dots, n_A | \gamma_{ij} = h_p\}$, and collect all such sets in the nested list $\mathcal{R} = \{r_{p_j} | p \in \{1, \ldots, P\}, j \in \{1, \ldots, n_B\} \}$. We compute the number of records in $A$ that share agreement pattern $p$ with record $B_j$, 
\begin{align}\label{eqn:N}
N_{p_j} = |r_{p_j}| = \sum_{i=1}^{n_A} I(\gamma_{ij} = h_p).
\end{align}
We collect these counts in $\mathcal{N} = \{N_{p_j} |p \in 1, \ldots, P, j \in 1, \ldots, n_B \}$.

The set $\tilde{\gamma} = \{\mathcal{P}, \mathcal{R}, \mathcal{N}\}$ fully characterizes the comparison matrix $\gamma$ for purposes of writing the likelihood function for \texttt{fabl}. To see this, we employ the condensed notation
\begin{subequations}
	\begin{align}
		m_p =  \textcolor{red}{p(\Gamma_{ij} = h_p|Z_j = i)} = \prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}\label{eqn:m_p} \\
		u_p =  \textcolor{red}{p(\Gamma_{ij} = h_p|Z_j \neq i)} = \prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}\label{eqn:u_p}
	\end{align}
\end{subequations}
to express the probability that records $A_i$ and $B_j$ form agreement pattern $p$ given that they are a match, in (\ref{eqn:m_p}), and not a match, in (\ref{eqn:u_p}). Viewed through the perspective of agreement patterns, the likelihood in (\ref{eqn:likelihood}) \textcolor{teal}{can be expressed as}
%is equivalent to

@@

Turning first to the \textcolor{teal}{full conditional updates}
%posterior updates 
for $\bm{m}$ and $\bm{u}$, we observe that these parameters only depend on the \textcolor{teal}{observed} data  through quantities that are calculated through $\mathcal{N}$ and $\mathcal{P}$. Let $n_p(\bm{Z}) = \sum_{j=1}^{n_B} I\left(\gamma_{Z_j, j} = h_p \right)$ be the number of matching record pairs with agreement pattern $p$, and $N_p = \sum_{j=1}^{n_B} N_{p_j}$ be the total occurrence of pattern $p$ in the data across all record pairs. Conditional on $\bm{Z}$, we can express the contribution to the likelihood in the full conditional distribution for $\bm{m}$ and $\bm{u}$ as
\begin{align}
	\mathcal{L}(\bm{m}, \bm{u} \mid \tilde{\gamma}, \bm{Z}, \pi)  &=  \prod_{p=1}^P m_p^{n_p(\bm{Z})}u_p^{N_p - n_p(\bm{Z})}  \label{likelihood_efficient_m_u}.
\end{align}
Additionally, let $\bm{\alpha_0} = (\alpha_{11}, \ldots, \alpha_{F L_F})$ and $\bm{\beta_0}= (\beta_{11}, \ldots, \beta_{F L_F})$ be a concatenated vectors of prior parameters for the $\bm{m}$ and $\bm{u}$ distributions respectively. The terms needed for the \textcolor{teal}{full conditional} updates for the $\bm{m}$ and $\bm{u}$ parameters are given by the appropriate components of the vectors
\begin{subequations}
	\begin{align}
		\bm{\alpha(Z}^{(s)}\bm{)} &= \bm{\alpha_0} + \sum_{p=1}^P n_p\left(\bm{Z}^{(s)}\right) \times e(h_p), \label{efficient_alpha} \\
		\bm{\beta(Z}^{(s)}\bm{)} &= \bm{\beta_0} + \sum_{p=1}^P \left(N_p - n_p\left(\bm{Z}^{(s)}\right)\right) \times e(h_p). \label{efficient_beta}
	\end{align}
\end{subequations}
%Specifically, the $\ell + \sum_{k=1}^{f-1} L_k$ component of (\ref{efficient_alpha}) and (\ref{efficient_beta}) provides the the posterior updates for level $\ell$ and field $f$ in (\ref{eqn:alpha_update}) and (\ref{eqn:beta_update}). However, the vectorized summations in (\ref{efficient_alpha}) and (\ref{efficient_beta}) can be computed more efficiently than summing over $n_A n_B$ record pairs for each field and agreement level.

@@

\textcolor{teal}{\textcolor{red}{We can} compute the vectorized summations in (\ref{efficient_alpha}) and (\ref{efficient_beta}) more efficiently than the previous updates in (\ref{eqn:alpha_update}) and (\ref{eqn:beta_update}), which require summing over $n_A n_B$ record pairs for each field and agreement level.}

\textcolor{teal}{Turning to sampling for $\pi,$ it remains the same as in (\ref{eqn:pi}), \textcolor{red}{so we do not repeat it.} }

We also use (\ref{likelihood_efficient}) to sample $\bm{Z}$ efficiently. For each pattern $p$, let $w_p = m_p/u_p$.  The contribution to the likelihood in the full conditional for $Z_j$ can be expressed as
\begin{align}
	\mathcal{L}(Z_j\mid  \tilde{\gamma}, \bm{m}, \bm{u}, \pi) = \prod_{p=1}^P u_p^{N_{p_j}} \prod_{i \in r_{p_j}} w_p^{I(Z_j = i)}. \label{likelihood_efficient_z}
\end{align}
This likelihood lends itself to markedly faster posterior computations. Sampling $Z_j$ from the full conditional provided in (\ref{eqn:z_full_conditional}) can become computationally expensive when $n_A$ is large. This is because sampling a value from $n_A$ options with unequal weights requires normalizing the weights to probabilities, which has a computational cost that scales with $n_A$. \textcolor{red}{We use two steps in sampling each $Z_j$.}
%To speed up computation, we break this sampling step into two.

We first sample among $P + 1$ options for the agreement pattern between $B_j$ and its potential link, according to

\begin{align}
	\label{eqn:gibbs1}
	p\left( \Gamma_{Z_j^{(s+1)},j} = h_p \mid \tilde{\gamma}, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)}\right) \propto
%	p\left( \Gamma_{Z_j^{(s+1)}, j} = h_p \mid \tilde{\gamma}, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)}\right) \propto
	\begin{cases} 
%		\frac{\pi^{(s+1)}N_{p_j}}{n_A}  w_{p}^{(s+1)},  & \gamma_{Z_j^{(s+1)}, j} = h_p; \\
                 \frac{\pi^{(s+1)}N_{p_j}}{n_A}  w_{p}^{(s+1)},  & h_p \in \mathcal{P}; \\
		1- \pi^{(s+1)} , &   \text{otherwise}. \\
	\end{cases}
\end{align}

Since all records in $A$ sharing the same agreement pattern with $B_j$ are equally likely, we then sample among candidate records uniformly using
\begin{align}
	\label{eqn:gibbs2}
%	P\left(Z_j^{(s+1)} = q \mid \gamma_{Z_j^{(s+1)}, j} = h_p, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)} \right) 
P\left(Z_j^{(s+1)} = q \mid \Gamma_j=h_p, \bm{m}^{(s+1)}, \bm{u}^{(s+1)}, \pi^{(s+1)} \right) 
	= 
	\begin{cases} 
		\frac{1}{N_{p_j}}, & q \in r_{p_j}; \\
		0, & \text{otherwise.} \\
	\end{cases}
\end{align}
